Hippocampal replay of experience at real-world speeds

### 2021-07-30 Article.tex
Revert "Revert "Update on Overleaf.""
This reverts commit 7bfb406708996248245fb81365e047950d24c3e1.

The brain has the remarkable ability to store and retrieve representations of past events, enabling memories of the past to influence future behavior. These memory processes depend critically on the hippocampus, where rapid plasticity during an experience is thought to drive the initial encoding of events \citep{EichenbaumConditioningConsciousRecollection2004}. Subsequently, hippocampal "replay" of stored representations during both slow-wave sleep and periods of immobility during waking is thought to contribute to the longer term storage and updating of event memories in distributed hippocampal-neocortical circuits \citep{Franklandorganizationrecentremote2005, Joohippocampalsharpwave2018, CarrHippocampalreplayawake2011}.

The canonical example of "replay" is seen in the rodent hippocampus, where cells preferentially fire at specific locations in an environment, and thus ensembles of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and thus are thought to "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow5-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay \citep{MichonPostlearningHippocampalReplay2019, KaeferReplayBehavioralSequences2020, ShinDynamicsAwakeHippocampalPrefrontal2019, DavidsonHippocampalReplayExtended2009, Tingleymethodsreactivationreplay2020}. One might therefore conclude that all memories that are stored or updated by SWRs correspond to sequences of locations. Our subjective experience, however, suggests that we can retrieve memories of individual locations without having to mentally traverse a long distance. Such memories would seem to be useful, in that they could encode the stimuli and values associated with a given place irrespective of the path used to get there. If the rodent hippocampus is capable of storing those sorts of memories, then one might expect to see SWRs where the neural activity corresponds not to a rapid trajectory through space, but instead to either a stable pattern associated with a single location or perhaps a brief pattern more similar to one that occurs during a real experience.

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{YuDistincthippocampalcorticalmemory2017}, although another report suggested that events that represent a single location are only seen in young animals \citep{FarooqEmergencepreconfiguredplastic2019}. Thus, while it would seem useful to replay memories associated with single locations or a small region of the environment, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approaches for detecting sequential replay involve fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, TingleyRoutingHippocampalRipples2020, BhattaraiDistincteffectsreward2020, ShinDynamicsAwakeHippocampalPrefrontal2019, DrieuNestedsequenceshippocampal2018, OlafsdottirTaskDemandsPredict2017, TangHippocampalPrefrontalReactivationLearning2017, CarrTransientSlowGamma2012, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data, where the shuffled version of the data represent an implicit definition of a "random" sequence.

@@

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the most probable location at each time bin \citep{KaeferReplayBehavioralSequences2020, CareyRewardrevaluationbiases2019, GrosmarkDiversityneuralfiring2016, WuHippocampalReplayCaptures2014, PfeifferHippocampalplacecellsequences2013}. For example, using this approach, \cite{PfeifferAutoassociativedynamicsgeneration2015} found awake replay events containing alternation between representations of a single location and sequential spatial trajectories. On the other hand, \cite{StellaHippocampalReactivationRandom2019} reported that replays during sleep are spatially continuous and follow Brownian diffusion dynamics. However, both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, making it hard to identify the source of the different conclusions.

An ideal approach to identifying and quantifying the dynamics of replay would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would provide a moment-by-moment estimate of position and dynamics that would not be dependent on the estimation of SWR start and end times and could rapidly adjust to changes in dynamics. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and would provide information about the certainty of the decoded estimates. It would be able to capture a range of movement dynamics from stationary or unchanging representations to trajectories that progress through space at constant or variable speeds to disorganized, spatially incoherent events. It would provide a robust statistical assessment of confidence for each dynamic. Finally, where assumptions are made, it would provide well-defined parameters whose values could be explored systematically to understand their influence on the results.

We therefore developed a state space model that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \citep{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing the spatial representations during SWRs as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to spiking data during SWR events from 10 rats, enabling a direct comparison to previous work.

@@

% Figure 1
\begin{figure}
\includegraphics[width=0.65\linewidth]{figures/Figure1/Figure1_final}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. For the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time (Stat. = Stationary, S-C-M = Stationary-Continuous-Mixture, Cont. = Continuous, F-C-M = Fragmented-Continuous-Mixture, Frag. = Fragmented), which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). Red dotted line in the right panel is the best fit line from the standard decoder using the Radon transform. \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.80 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{fig:1}
% Figure 1-Supplemental 1
\figsupp[The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms)]{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each panel shows the probability of each dynamic on simulated data example from \FIG{1} with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in \FIG{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in \FIG{1}F. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}{\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1-supplemental1_final}}\label{figsupp:fig1sup1}
\end{figure}

@@

We note here that the choice of any particular threshold for classifying the movement dynamic of a SWR is arbitrary, and that the power of our approach lies in part on the ability to assign a probability for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.80 was to use a threshold that corresponds to highly probable events and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold often leads to the conclusion that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verify that our central results hold with a higher threshold of 0.95. We do not know if downstream neurons can explicitly responds to these dynamics based on either threshold (there is probably no hard boundary between these dynamics), but the dynamics correspond to fundamentally different sets of speeds and can be used as a convenient tool in order to understand and summarize the SWR.

@@

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \citep{TrautmannAccurateEstimationNeural2019} and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly uses multiunit spikes and their spike waveform features to decode position without spike sorting (see Methods). Clusterless decoding has previously been used to successfully identify theta sequences and replay sequences in the hippocampus \citep{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (\FIG{2}B, D), both with 1D linearized position and 2D position (\FIGSUPP[2]{fig2sup1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of latent position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (\FIG{2}D vs C).

\subsection*{Most replays contain coherent spatial content}
After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}; position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal. Even including theta sequences and replay sequences, which can represent positions other than where the animal is, we found that the average median difference between actual and decoded location is small (7 cm median difference, 5-9 cm 95\% CI, for all times where the animal was moving greater than 4 cm/s, 5 fold cross validation). This level of accuracy is just above the size of our 3 cm position bins, and is comparable to that of other studies \citep{DavidsonHippocampalReplayExtended2009, ShinDynamicsAwakeHippocampalPrefrontal2019, FarooqEmergencepreconfiguredplastic2019}. Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during movement, demonstrating the power of our algorithm. For each animal, we also verified that the population spiking rate was relatively consistent over all positions on the track, suggesting that we have comparable sampling of all locations in the environment (\FIGSUPP[2]{fig2sup3}).

@@

% Figure 3
\begin{figure}
\includegraphics[width=0.70\linewidth]{figures/Figure3/Figure3_final}
\caption{A-F. Examples of SWRs with non-constant speed trajectories. Figure conventions are the same as in \FIG{2}. Filtered SWR ripple (150-250 Hz) trace from the tetrode with the maximum amplitude displayed above each example. \textbf{(A)} A SWR where the decoded position starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A SWR where the decoded position persistently stays at the choice point (open circle) while the animal remains at the left well. \textbf{(C)} A SWR where the decoded position begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A SWR where the decoded position begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} A SWR where the decoded position begins in the left arm and persists at the end of the center arm. \textbf{(F)} A SWR where the decoded position starts in the left arm toward the choice point, jumps to the right arm and proceeds back toward the choice point. \textbf{(G)} Classification of SWRs from multiple animals and datasets. Each dot represents the percentage of SWRs for each day for that animal. A SWR is included in the numerator of the percentage if at any time it includes the classifications listed below the column. The denominator is listed in the x-axis label.
}
\label{fig:3}

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% (22,170 of 25,844) of all SWRs include spatially coherent structure, and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs include any times with spatially incoherent structure (\FIG{3}G). To further validate this result, we performed a shuffling procedure that, in contrast to our previous shuffle, preserves the local correlations between spikes but reduces global spatial structure. This controls for dynamics that may have been induced by bursts or other local features of the spike train while randomizing the position to spike relationship. To do this, we randomized the order of runs (from one reward well to another) and then circularly permuted the resulting segments of data across all tetrodes uniformly. Using this shuffle, we found that there was still less spatially coherent and more spatially incoherent content compared to the real data (recording session \#1, spatially coherent: 99\% vs. 71\%, p=0.02, real vs. shuffled; spatially incoherent: 22\% vs. 61\%, p=0.02, real vs. shuffled; recording session \#2, spatially coherent: 95\% vs. 63\%, p=0.02, real vs. shuffled; spatially incoherent: 16\% vs. 79\%, p=0.02, real vs. shuffled; \FIGSUPP[3]{fig3sup3}).

@@

% Figure 4
\begin{figure}
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_final}
\caption{Validation of classification using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. In each column: top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. Middle panel: Posterior distribution of estimated linear position over time. Magenta dashed line indicates animal's position. Bottom panel: HPD region size---the integral of the position bins with the top 95\% probable values. \textbf{(E)} Average 95\% HPD region size for each dynamic category. Median 24 cm for spatially coherent vs. median 238 cm for spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U}
\label{fig:4}

@@

We then carried out a series of control analyses to determine whether our results could be due to differences in spiking statistics across dynamics or interneuron-specific activity patterns. We first calculated the multiunit spiking rates and found that these were similar across the different dynamics. (\FIG{5}F). This indicates that all dynamics, including stationary and stationary-continuous mixtures, were driven by comparable levels of sustained spiking information and could not be explained by the absence of spiking. Further corroborating this, we found that when we re-ran our analysis using periods of high multiunit activity instead of SWRs, we still found a larger proportion of stationary and stationary-continuous dynamics relative to the continuous dynamics \FIGSUPP[5]{fig5sup3}F).

>>>>

The brain has the remarkable ability to store and retrieve representations of past events, enabling memories of the past to influence future behavior. These memory processes depend critically on the hippocampus, where rapid plasticity during an experience is thought to drive the initial encoding of events \citep{EichenbaumConditioningConsciousRecollection2004}. Subsequently, hippocampal "replay" of stored representations during both slow-wave sleep and periods of immobility during waking is thought to contribute to the longer term storage and updating of event memories in distributed hippocampal-neocortical circuits \citep{Franklandorganizationrecentremote2005, CarrHippocampalreplayawake2011, Joohippocampalsharpwave2018}.

The canonical example of "replay" is seen in the rodent, where hippocampal cells preferentially fire at specific locations in an environment, and thus ensembles of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and thus are thought to "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow5-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay \citep{DavidsonHippocampalReplayExtended2009, MichonPostlearningHippocampalReplay2019,
ShinDynamicsAwakeHippocampalPrefrontal2019, KaeferReplayBehavioralSequences2020, Tingleymethodsreactivationreplay2020}. One might therefore conclude that all memories that are stored or updated by SWRs correspond to sequences of locations. Our subjective experience, however, suggests that we can retrieve memories of individual locations without having to mentally traverse a long distance. Such memories would seem to be useful, in that they could encode the stimuli and values associated with a given place irrespective of the path used to get there. If the rodent hippocampus is capable of storing those sorts of memories, then one might expect to see SWRs where the neural activity corresponds not to a rapid trajectory through space, but instead to either a stable pattern associated with a single location or perhaps a brief pattern more similar to one that occurs during a real experience.

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{YuDistincthippocampalcorticalmemory2017}, although another report suggested that events that represent a single location are only seen in young animals \citep{FarooqEmergencepreconfiguredplastic2019}.  Other studies have reported activity that, when combined across an entire event, seems to correspond roughly to a single location, although the dynamics of these events is typically not investigated (REF Dupret, Csicvari papers with cheeseboard task). Thus, while it would seem useful to replay memories associated with single locations or a small region of the environment, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approaches for detecting sequential replay involve fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009, CarrTransientSlowGamma2012, OlafsdottirTaskDemandsPredict2017, TangHippocampalPrefrontalReactivationLearning2017, DrieuNestedsequenceshippocampal2018,  ShinDynamicsAwakeHippocampalPrefrontal2019, TingleyRoutingHippocampalRipples2020, BhattaraiDistincteffectsreward2020}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data, where the shuffled version of the data represent an implicit definition of a "random" sequence.

@@

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the most probable location at each time bin \citep{PfeifferHippocampalplacecellsequences2013, WuHippocampalReplayCaptures2014, GrosmarkDiversityneuralfiring2016, CareyRewardrevaluationbiases2019,  KaeferReplayBehavioralSequences2020}. For example, using this approach, \cite{PfeifferAutoassociativedynamicsgeneration2015} found awake replay events containing alternation between representations of a single location and sequential spatial trajectories. On the other hand, \cite{StellaHippocampalReactivationRandom2019} reported that replays during sleep are spatially continuous and follow Brownian diffusion dynamics. However, both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, making it hard to identify the source of the different conclusions.

An ideal approach to identifying and quantifying the dynamics of replay would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would provide a moment-by-moment estimate of position and dynamics that would not be dependent on the estimation of SWR start and end times and could rapidly adjust to changes in dynamics. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and would provide information about the certainty of the decoded estimates. It would be able to capture a range of movement dynamics including stationary or unchanging representations, trajectories that progress through space at constant or variable speeds, and disorganized, spatially incoherent events. It would provide a robust statistical assessment of confidence for each dynamic. Finally, where assumptions are made, it would provide well-defined parameters whose values could be explored systematically to understand their influence on the results.

We developed a state space model that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \citep{DengRapidclassificationhippocampal2016, MaboudiUncoveringtemporalstructure2018} by characterizing the spatial representations during SWRs as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to spiking data during SWR events from 10 rats, enabling a direct comparison to previous work.

@@

% Figure 1
\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/Figure1/Figure1_final}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. For the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time (Stat. = Stationary, S-C-M = Stationary-Continuous-Mixture, Cont. = Continuous, F-C-M = Fragmented-Continuous-Mixture, Frag. = Fragmented), which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). Red dotted line in the right panel is the best fit line from the standard decoder using the Radon transform. \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.80 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{fig:1}
% Figure 1-Supplemental 1
\figsupp[The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms)]{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each panel shows the probability of each dynamic on simulated data example from \FIG{1} with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in \FIG{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in \FIG{1}F. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}{\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1-supplemental1_final}}\label{figsupp:fig1sup1}
\end{figure}

@@

We note here that the choice of any particular threshold for classifying the movement dynamic of a SWR is arbitrary, and that the power of our approach lies in part on the ability to assign a probability for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.80 was to use a threshold that corresponds to highly probable events and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold often leads to the conclusion that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verify that our central results hold with a higher threshold of 0.95. We do not know if downstream neurons can explicitly responds to these dynamics based on either threshold, and there is probably no hard boundary between these dynamics, but our approach makes it possible to ask that question in a systematic way.

@@

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \citep{TrautmannAccurateEstimationNeural2019} and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly uses multiunit spikes and their spike waveform features to decode position without spike sorting (see Methods). Clusterless decoding has previously been used to successfully identify theta sequences and replay sequences in the hippocampus \citep{ChenTransductiveneuraldecoding2012, KloostermanBayesiandecodingusing2014, DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (\FIG{2}B, D), both with 1D linearized position and 2D position (\FIGSUPP[2]{fig2sup1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of latent position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (\FIG{2}D vs C).

\subsection*{Most replays contain coherent spatial content}
After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, CarrTransientSlowGamma2012, KayConstantSubsecondCycling2020}; position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal. We found that the average median difference between actual and decoded location is small (7 cm median difference, 5-9 cm 95\% CI, for all times where the animal was moving greater than 4 cm/s, 5 fold cross validation). This level of accuracy is comparable to that of other studies \citep{DavidsonHippocampalReplayExtended2009, ShinDynamicsAwakeHippocampalPrefrontal2019, FarooqEmergencepreconfiguredplastic2019}, and reflects in part the presence of "theta sequences" [REF] where position deviates behind and ahead of the animal across individual theta cycles.  Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during movement, demonstrating the power of our algorithm. For each animal, we also verified that the population spiking rate was relatively consistent over all positions on the track, suggesting that we have comparable sampling of all locations in the environment (\FIGSUPP[2]{fig2sup3}).

@@

% Figure 3
\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/Figure3/Figure3_final}
\caption{A-F. Examples of SWRs with non-constant speed trajectories. Figure conventions are the same as in \FIG{2}. Filtered SWR ripple (150-250 Hz) trace from the tetrode with the maximum amplitude displayed above each example. \textbf{(A)} A SWR where the decoded position starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A SWR where the decoded position persistently stays at the choice point (open circle) while the animal remains at the left well. \textbf{(C)} A SWR where the decoded position begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A SWR where the decoded position begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} A SWR where the decoded position begins in the left arm and persists at the end of the center arm. \textbf{(F)} A SWR where the decoded position starts in the left arm toward the choice point, jumps to the right arm and proceeds back toward the choice point. \textbf{(G)} Classification of SWRs from multiple animals and datasets. Each dot represents the percentage of SWRs for each day for that animal. A SWR is included in the numerator of the percentage if at any time it includes the classifications listed below the column. The denominator is listed in the x-axis label.
}
\label{fig:3}

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{FosterReversereplaybehavioural2006, KarlssonAwakereplayremote2009, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% (22,170 of 25,844) of all SWRs include spatially coherent structure, and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs include any times with spatially incoherent structure (\FIG{3}G). To further validate this result, we performed a shuffling procedure that, in contrast to our previous shuffle, preserves the local correlations between spikes but reduces global spatial structure. This controls for dynamics that may have been induced by bursts or other local features of the spike train while randomizing the position to spike relationship. To do this, we randomized the order of runs (from one reward well to another) and then circularly permuted the resulting segments of data across all tetrodes uniformly. Using this shuffle, we found that there was still less spatially coherent and more spatially incoherent content compared to the real data (recording session \#1, spatially coherent: 99\% vs. 71\%, p=0.02, real vs. shuffled; spatially incoherent: 22\% vs. 61\%, p=0.02, real vs. shuffled; recording session \#2, spatially coherent: 95\% vs. 63\%, p=0.02, real vs. shuffled; spatially incoherent: 16\% vs. 79\%, p=0.02, real vs. shuffled; \FIGSUPP[3]{fig3sup3}).

@@

% Figure 4
\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/Figure4/Figure4_final}
\caption{Validation of classification using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. In each column: top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. Middle panel: Posterior distribution of estimated linear position over time. Magenta dashed line indicates animal's position. Bottom panel: HPD region size---the integral of the position bins with the top 95\% probable values. \textbf{(E)} Average 95\% HPD region size for each dynamic category. Median 24 cm for spatially coherent vs. median 238 cm for spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U}
\label{fig:4}

@@

We then carried out a series of control analyses to determine whether our results could be due to differences in spiking statistics across dynamics or interneuron-specific activity patterns. We first calculated the multiunit spiking rates and found that these were similar across the different dynamics. (\FIG{5}F). This indicates that all dynamics, including stationary and stationary-continuous mixtures, were driven by comparable levels of sustained spiking information and could not be explained by the absence of spiking. Further corroborating this, we found that when we re-ran our analysis using periods of high multiunit activity instead of SWRs to identify events, we still found a larger proportion of stationary and stationary-continuous dynamics relative to the continuous dynamics \FIGSUPP[5]{fig5sup3}F).
